### 2013-10-17 src2/thesis.tex
apa

Paper E, was the first I wrote with Josef Svenningsson who had been thinking about an interesting
variation of counting sort and wanted to try implementing it. Together we made the required changes
to Obsidian. Me and Josef are equal partners in this paper and got much valuable advice from 
Mary Sheeran.

@@

@@

The patterns are powerful abstractions, but it is also important not to be locked into 
a single implementation of each pattern. This could have negative performance impacts. 
In ArBB, for example, the patterns used are specialised to the current hardware during JIT. 
Languages like OpenCL and CUDA do not provide abstractions of these patterns. The programmer 
needs to implement them using lower level features making CUDA and OpenCL code less composable 
and harder to reuse. As an example, look at the slides from NVIDIA in reference~\citet{reduction} 
where a reduction primitive is implemented in CUDA. First a naive reduction kernel is implement 
and it is refined over many steps, taking many characteristics of the GPU into mind. What is 
is that says that we do not need to redo this exploration with each new GPU generation 
and discover a new optimal solution?

@@

%\noindent\emph{Obsidian, Embedded languages}

@@

\subsubsection{Pull arrays}

\subsubsection{Push arrays}

%\subsubsection{Fusion for free}

@@

\subsection{Background} 
\noindent\emph{What does ArBB do, what are the operations and datastructures it support}

\noindent\emph{What does EmbArBB do}

@@

@@

\subsection{The results}

@@

@@

\subsubsection{Retargetable parallel programming}

@@

>>>>

Paper E, was the first I wrote with Josef Svenningsson who had been thinking about an interesting
variation of counting sort and wanted to implement it. Together we added atomic operations to 
the deep embedding and I modified the Obsidian code generator to accomodate them. Sections 
1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
the benchmarking. Section 8 is joint work.
%Me and Josef are equal partners in this paper and got much valuable advice from 
%Mary Sheeran.

@@

%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a fuctioning 
%code generating backend u

@@

The patterns are powerful abstractions, but it is also important not to be locked into 
a single implementation of each pattern. This could have negative performance impacts. 
In ArBB, for example, the patterns used are specialised to the current hardware during JIT 
compilation. Languages like OpenCL and CUDA do not provide abstractions of these patterns. 
The programmer needs to implement them using lower level features making CUDA and OpenCL 
code less composable and harder to reuse. As an example, look at the slides from NVIDIA in 
reference~\citet{reduction} where a reduction primitive is implemented in CUDA. First a 
naive reduction kernel is implement and it is refined over many steps, taking many 
characteristics of the GPU into mind. The end result is a reduction kernel that performs 
very well on the chosen GPU. But as GPUs evolve the design space may need to be explored again. 
%What is 
%is that says that we do not need to redo this exploration with each new GPU generation 
%and discover a new optimal solution?

@@

@@

\subsubsection{Pull arrays}
\emph{fusion for free}

\subsubsection{Push arrays}
\emph{fusion for free}

@@

Intel ArBB~\citet{ARBB2011} is a system for high-level data-parallel programming with the 
ability to generate code for a variety of different hardware confgurations. It is implemented 
as an embedded language in C++. One motivation for ArBB that is mentioned in~\citet{ARBB2011} is 
that while high-performance computing specialist are highly competent at implementing 
kernels using low-level programming techniques, mainstream developers are not. ArBB offers 
a more composable way to write programs that make use of core and vector (SIMD) parallelism and 
doing this while using a familiar language, C++.

One of the greatest strengths of ArBB, as I see it, is that it also comes with a low-level C 
interface. The main purpose of this C interface is to make it easier to use the ArBB system 
from other languages. Many languages have foreign function interfaces that are geared towards C.
This gives ArBB a level of language independence on top of its cross platform abilities.

ArBB is based on a set of parallel primitives (or structures) on dense 1,2 and 3 dimensional 
arrays and nested arrays: 
\begin{itemize} 
\item {\bf Reductions}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
\item {\bf Scans}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
\item {\bf Sorting}      {\tt sort} 
\item {\bf Permutations} {\tt gather}, {\tt scatter}, {\tt reverse} ...
\item {\bf Sequential controlflow} {\tt \_for}, {\tt \_while} 
\item {\bf Map, zipWith, stencil}  {\tt map} 
\end{itemize}

These operations (and many more, there are about 120 operations in total) can be used 
by the programmer when writing programs using ArBB. Note that there are sequential control 
operations with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
language and calling the ArBB functions does not actually execute them, it just ads a node to 
some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
This is exactly the way these things work in Haskell but it may be even more surprising to a
C++ programmer. Before an ArBB function can be computed it needs to be captured. For this 
ArBB has a {\tt call} function that is used to run functions using ArBB functionality. The 
first time {\tt call} is used on a function pointer that function is executed and the ArBB 
functions used in it builds an AST (we say the function has been captured). The AST is then 
compiled and specialised for the hardware available. Any subsequent calls of a function using 
ArBB functionality does not lead to a recompilation, the first call caches this code.

The first step of our work with embedding Intel ArBB functionality was to implement very direct 
and low-level bindings to the C interface. This provides a Haskell function (in the IO Monad) 
for each of the functions in the ArBB C interface. More details and results of this work 
can be found in paper G in section~\ref{sec:paperG}.

%\noindent\emph{What does ArBB do, what are the operations and datastructures it support}

%\noindent\emph{What does EmbArBB do}

@@

The ArBB Haskell bindings provide a very rudimentary interface to ArBB functionality and 
is not useful for application implementation. A higher-level interface to ArBB to offer 
the Haskell programmer is needed. Our first attempt was to implement an Accelerate backend 
using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
more general language and there was an API missmatch between them. These problems 
are mentioned in in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
to the Haskell in a way that is useful and simpler to implement, we start working on EmbArBB.

With EmbArBB we improve the interfacing with Haskell by implementing a deeply embedded language. 
Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings creating 
function objects that be run.

@@

@@

%\subsection{The results}

\subsection{Current status of ArBB and EmbArBB}

Since our work with ArBB, Intel has unfortunately retired the ArBB system. This 
leaves EmbArBB without a functioning code generating backend.

@@

\subsection{Retargetable parallel programming}

@@

\subsection{Contributions}

This thesis contributes to the follwoing areas:

\begin{itemize} 
\item Embedded language design and implementation. 
\item General Purpose programming using a GPU (GPGPU).
\item High level data-parallel programming across devices.
\end{itemize}

\subsubsection{Embedded language design and implementation}

\subsubsection{General Purpose programming using a GPU}

\subsubsection{High level data-parallel programming across devices}
